Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-03-31 abstract.tex
proofreading + images etc.

\begin{abstract}
Learning the dependency structure of a multivariate distribution from
the observational data is an important task since it allows us to
speculate about the underlying causal mechanisms that induce
dependencies. This learning task can be approached as a model
selection problem, but recent studies have revealed that the popular
Bayesian model selection criterion is not satisfactory. We review some
of the information theoretic alternatives and introduce a new one
called a quotient normalized maximum likelihood criterion.
\end{abstract}

>>>>

\begin{abstract}
Learning the dependency structure of a multivariate distribution from
observational data is an important task since it allows us to
speculate about the underlying causal mechanisms that induce
dependencies. This learning task can be approached as a model
selection problem, but recent studies have revealed that the popular
Bayesian model selection criteria are not satisfactory. We review some
of the information theoretic alternatives and introduce a new criterion
called the quotient normalized maximum likelihood criterion.
\end{abstract}

### 2017-03-31 bn.tex
proofreading + images etc.

Bayesian networks are a general way to describe the dependencies
between the components of an $n$\nobreakdash-dimensional discrete data
vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
Despite denoting the values with small integers, the model will treat
the components of $X$ categorical.

@@

Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there are a directed edges to the variable
$X_{i}$.

Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
$G_{i}(x)$. It is customary to enumerate all the possible
sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
$P(G_{i}(x)=1)=1$ for all vectors $x$.

@@

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
\end{eqnarray}
In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
in a closed form~\cite{Bunt91, Heck95}.

@@

Finding the satisfactory Dirichlet hyperparameters for the Bayesian
mixture above has, however, turned out to be problematic. Early on,
one of the desiderata for a good model selection criterion was that it
would yield equal scores for essentially equivalent
models~\cite{Verm90}.  For example, the score for the structure
$X_1\rightarrow X_2$ should be the same as the score for the model
$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
that variables $X_1$ and $X_2$ are statistically dependent on each
other.  It can be shown~\cite{Heck95} that to achieve this, not all
the hyperparameters $\alpha$ are possible and for practical reasons
Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
it soon turned out that BDeu score was very sensitive to the selection
of this hyperparameter~\cite{cosco.uai07} and that for small sample
sizes this method detects spurious correlations~\cite{Steck08} leading
to models with suspiciously many parameters.

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu of score and showed that in certain settings
BDeu has tendency to add more and more parent variables for a child
node even though the empirical conditional entropy of the child given
the parents has already reached zero. More in detail, assume that in
our data $D$, the values of $X_i$ are completely determined by
variables in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we
can further find one or more variables, denoted by $Y$, whose values
are determined completely by the variables in $Z$, then BDeu will
prefer the set $Z\cup Y$ over mere $Z$ as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates regularity in model
selection as the more complex model is preferred over a simpler one
even though it does not fit to data any better. The phenomena seems to
stem from the way the hyperparameters for the Dirichlet distribution
are chosen in BDeu as the BD score based on the Jeffreys prior does
not suffer from this anomaly.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by adopting the
NML-code to the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values.  While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

However, equation (\ref{eqn:szp1}) is derived only for the case
where $r$ is constant and $N$ grows. While with
fNML it is typical that $N$ is large compared to $r$, an
approximation for all ranges of $N$ and $r$ derived by
Szpankowski and Weinberger~\cite{Szpankowski2012} can also be used:
\begin{eqnarray}
\label{eqn:szp2}
    reg(N, r) & \approx & N\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha}
                - \frac{1}{C_\alpha}\right)\nonumber \\
    && - \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
\end{eqnarray}
where $\alpha = \frac{r}{N}$ and
\begin{equation}
    C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}.
\end{equation}
These approximations are compared in Table \ref{tbl:regrets}
to the exact regret for various values of $N$ and $r$.
For a constant $N$, equation (\ref{eqn:szp1}) provides
a progressively worse approximation as $r$ grows. Equation
(\ref{eqn:szp2}) on the other hand is a good approximation of
the regret regardless of the ratio of $N$ and $r$.

fNML solves the parameter sensitivity problem and yields predictive
models superior to BDeu.  However, the criterion does not satisfy the
property of giving the same score for the models that correspond to
the same dependence statements. Furthermore, the learned structures
still sometimes appear surprisingly complicated.

>>>>

Bayesian networks are a general way to describe the dependencies
between the components of an $n$\nobreakdash-dimensional discrete data
vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
Despite denoting the values with small integers, the model will treat
the components of $X$ as categorical variables.

@@

A Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there are a directed edges to the variable
$X_{i}$.

Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
of $x$ that consists of the values of the parents of $X_{i}$ in $x$ by
$G_{i}(x)$. It is customary to enumerate all the possible
sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
$P(G_{i}(x)=1)=1$ for all vectors $x$.

@@

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}
P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
\end{eqnarray}
In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
in a closed form~\cite{Bunt91, Heck95}.

@@

Finding satisfactory Dirichlet hyperparameters for the Bayesian
mixture above has, however, turned out to be problematic. Early on,
one of the desiderata for a good model selection criterion was that it
is score-equivalent, i.e., it would yield equal scores for
essentially equivalent models~\cite{Verm90}.  For example, the score
for the structure $X_1\rightarrow X_2$ should be the same as the score
for the model $X_2 \rightarrow X_1$ since they both correspond to the
hypothesis that variables $X_1$ and $X_2$ are statistically dependent
on each other.  It can be shown~\cite{Heck95} that to achieve this,
not all the hyperparameters $\alpha$ are possible and for practical
reasons Buntine~\cite{Bunt91} suggested a so called BDeu score with
just one hyperparameter $\alpha\in R_{++}$ so that
$\theta_{ij\cdot}\sim Dir(\frac{\alpha}{q_i
  r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However, it soon turned out
that the BDeu score was very sensitive to the selection of this
hyperparameter~\cite{cosco.uai07} and that for small sample sizes this
method detects spurious correlations~\cite{Steck08} leading to models
with suspiciously many parameters.

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates regularity in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score-equivalent.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML-code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

However, equation (\ref{eqn:szp1}) is derived only for the case
where $r$ is constant and $N$ grows. While with
fNML it is typical that $N$ is large compared to $r$, an
approximation for all ranges of $N$ and $r$ derived by
Szpankowski and Weinberger~\cite{Szpankowski2012} can also be used:
\begin{eqnarray}
\label{eqn:szp2}
    reg(N, r) & \approx & N\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha}
                - \frac{1}{C_\alpha}\right)\nonumber \\
    && - \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
\end{eqnarray}
where $\alpha = \frac{r}{N}$ and
\begin{equation}
    C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}.
\end{equation}
These approximations are compared in Table \ref{tbl:regrets} to the
exact regret for various values of $N$ and $r$.  For a constant $N$,
equation (\ref{eqn:szp1}) provides a progressively worse approximation
as $r$ grows. Equation (\ref{eqn:szp2}) on the other hand is a good
approximation of the regret regardless of the ratio of $N$ and $r$.
In our experiments, we will use this approximation for 
implementation of the qNML criterion.

fNML solves the parameter sensitivity problem and yields predictive
models superior to BDeu.  However, the criterion does not satisfy the
property of giving the same score for models that correspond to
the same dependence statements. Furthermore, the learned structures
are often rather complex which hampers their interpretation.

### 2017-03-31 experiments.tex
proofreading + images etc.

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and run 1000 train and test experiments
for 20 different sample sizes for all of them. The training was
conducted using a dynamic programming based exact structure
learning\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

Figure~\ref{fig:bcmean} shows an example of the behavior of the
studied selection criteria.  For small sample sizes fNML and qNML
yield equally accurate predictions while BDeu requires more samples to
obtain competitive performance. The curves are mean curves from 1000
independent train and test splits.

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
\caption{Predictive log-loss in a breast cancer data as a function of
  sample size for different model selection criteria.}
\label{fig:bcmean}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
\caption{Number of parameters in a breast cancer model as a function
  of sample size for different model selection criteria.}
\label{fig:bcnpmean}
\end{figure}

\begin{table}
\centering
\begin{tabular}{crrrr}
\toprule
           Data &     N &   BDeu &            fNML &            qNML \\
\midrule
           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
 HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
 HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
\bottomrule
\end{tabular}
\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
\label{tbl:preds}
	\end{table}

Table~\ref{tbl:preds} features predictive losses for the small sample
sizes.  (The results for large sample sizes converge for all sensible
criteria.)  As we can see, fNML still usually (12/20) performs best
but qNML is not much worse while BDeu may sometimes perform much
worse.

Looking at the number of parameters for the same datasets and sample
sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
yields simplest models, and sometimes, like in the BreastCancer data,
the differences are of orders of magnitude.

\begin{table}
\centering
\begin{tabular}{crrrr}
\toprule
           Data &     N &          BDeu &  fNML &          qNML \\
\midrule
           Iris &    16 &            36 &    33 &   \textbf{28} \\
  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
           Wine &    27 &         11559 &   780 &  \textbf{214} \\
 HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
 HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
          Liver &    54 &            40 &    60 &   \textbf{24} \\
          Glass &    55 &          1488 &   572 &  \textbf{100} \\
        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
        Balance &    96 &   \textbf{20} &   237 &           168 \\
    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
          Yeast &   225 &   \textbf{78} &   307 &            80 \\
        Abalone &   418 &            90 &   156 &   \textbf{62} \\
     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
\bottomrule
\end{tabular}
\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
\label{tbl:nofparams}
\end{table}

>>>>

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art4_mean.pdf}
\caption{Finding generating models of 4 arcs.}
\label{fig:4arcs}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
\caption{Finding generating models of 7 arcs.}
\label{fig:4arcs}
\end{figure}

We empirically compare the capacity of qNML to that of BIC, BDeu and
fNML in identifying the data generating structures, and producing
models that are predictive and parsimonious.  It seems that none of
the criteria uniformly outperform the others in all these desirable
aspects of model selection criteria.

\subsection{FINDING GENERATING STRUCTURE}

We first compare the ability of the different scoring criteria to
discover the structure that generated the data. For this purpose we
create 100 random 5-node Bayesian network structures with 4
edges and another 100 structures with 7 edges.  The variables were
randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
each network, we generated parameters by two different schemes. The
first scheme exactly matched the assumptions of the BDeu score with
$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
was to generate the parameters independently from a Dirichlet
distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
that is the Jeffreys' prior for the multinomial model. This distribution
was selected instead of the uniform distribution in order to make the
generating structure more identifiable.  For each network (structure +
parameters), we generated 100 data sets of 1000 data vectors, and
studied how different scoring criteria ranked the structure of the
generating network among all the 5-node networks as a function of
(sub)sample size.

Not surprisingly, the results indicate that when the parameter generation
mechanism matches the assumptions of the BDeu-score, BDeu usually
also ranks the generating structure higher than the other scores
(Figure X).  On the other hand, when the parameters are drawn from the
Jeffreys' prior, the fNML score appears to rank the generating structure
highest. This too is not surprising, since the NML distribution for
multinomial data is known to closely match the marginal distribution of
the data when the Jefrreys' prior is used.

The underfitting tendency of BIC can also be clearly detected both for
relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
ranking ability the qNML criterion appears to be between fNML and BDeu
in 3 out of 4 settings which is good considering that one of these
criteria is always the winner. For the dense networks with BDeu prior,
qNML appears to perform worse than BDeu and fNML but still much
better than BIC.

\subsection{PREDICTION AND PARSIMONY}

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
for all of them. In each experiment, random 10\% of the original
dataset was selected as training data and the rest 90\% as test data.
The training was conducted using a dynamic programming based exact structure
learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

When predicting with structures learned by the BDeu score, we used the
Bayesian expected parameter values $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter free, for structures learned by the other model
selection criteria, we used the predictive conditional NML
parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
\caption{Number of parameters in a breast cancer model as a function
  of sample size for different model selection criteria.}
\label{fig:bcnpmean}
\end{figure}

\begin{table}
\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
\label{tbl:preds}
\begin{center}
 \begin{tabular}{crrrrr}
    Data &     N &              BDeu &               BIC &               fNML &               qNML \\
\midrule
    Iris &    15 &     \textit{3.61} &              3.56 &   \underline{3.50} &      \textbf{3.47} \\
 PostOpe &    18 &    \textit{10.52} &     \textbf{7.45} &   \underline{8.35} &               8.39 \\
   Ecoli &    34 &     \textit{6.27} &              6.16 &      \textbf{5.56} &   \underline{5.63} \\
   Liver &    35 &     \textit{4.22} &     \textbf{4.01} &               4.10 &   \underline{4.03} \\
    Wine &    36 &    \textit{16.41} &    \textbf{11.27} &              12.07 &  \underline{11.86} \\
   Glass &    44 &     \textit{7.48} &              6.65 &      \textbf{6.37} &   \underline{6.43} \\
 Thyroid &    44 &              2.92 &     \textit{2.93} &      \textbf{2.81} &   \underline{2.83} \\
 HeartSt &    54 &    \textit{14.37} &    \textbf{10.96} &              11.98 &  \underline{11.65} \\
 BreastC &    58 &    \textit{11.21} &     \textbf{9.94} &  \underline{10.34} &              10.37 \\
 HeartHu &    60 &     \textit{9.21} &     \textbf{8.38} &               8.80 &   \underline{8.61} \\
 HeartCl &    62 &    \textit{14.88} &    \textbf{11.76} &              12.48 &  \underline{12.18} \\
 BcWisco &    70 &     \textit{6.41} &     \textbf{5.09} &               5.29 &   \underline{5.28} \\
 Diabete &    77 &     \textit{5.11} &  \underline{4.99} &               5.09 &      \textbf{4.98} \\
 TicTacT &    96 &             10.99 &    \textbf{10.35} &  \underline{10.72} &     \textit{11.04} \\
 Balance &   126 &              7.45 &  \underline{7.42} &      \textbf{7.27} &      \textit{7.73} \\
   Yeast &   149 &              5.67 &     \textit{5.80} &      \textbf{5.48} &   \underline{5.49} \\
 Abalone &   418 &  \underline{3.93} &     \textit{4.00} &      \textbf{3.88} &               3.96 \\
 PageBlo &   548 &  \underline{2.36} &     \textit{2.40} &      \textbf{2.32} &               2.38 \\
 Shuttle &  5800 &  \underline{1.69} &     \textit{1.72} &      \textbf{1.69} &               1.71 \\
   Adult &  6513 &             10.16 &    \textit{10.25} &     \textbf{10.05} &  \underline{10.08} \\
\end{tabular}
\end{center}
\end{table}

Table~\ref{tbl:preds} features predictive losses for the small sample
sizes.  (The results for large sample sizes converge for all sensible
criteria.) BIC's bias for simplicity makes it often win (written bold
in the table) with small sample sizes, but it performs worst (in
italics) for the large sample sizes. the qNML seems to achieve a lot
of runner-up positions (underlined) making it a rather safe choice.

@@

Looking at the number of parents for our 20 datasets and sample
sizes (Table~\ref{tbl:nofparams}) again features BIC's preference for
simple models. qNML usually (18/20) beats fNML that performs worst in
14 out of 20 data sets.

\begin{table}
  \caption{Average number of parents per node 
    for different model selection criteria in 20 different datasets.}
\label{tbl:nofparams}
\begin{center}
\begin{tabular}{crrrrr}
    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
\midrule
    Iris &    15 &              0.84 &     \textbf{0.63} &  \textit{0.87} &  \underline{0.82} \\
 PostOpe &    18 &              1.35 &     \textbf{0.11} &  \textit{1.66} &  \underline{1.29} \\
   Ecoli &    34 &              0.83 &     \textbf{0.30} &  \textit{0.96} &  \underline{0.77} \\
   Liver &    35 &              0.60 &     \textbf{0.10} &  \textit{0.88} &  \underline{0.43} \\
    Wine &    36 &     \textit{2.11} &     \textbf{0.75} &           1.86 &  \underline{1.44} \\
   Glass &    44 &              1.32 &     \textbf{0.38} &  \textit{1.41} &  \underline{0.85} \\
 Thyroid &    44 &              0.84 &     \textbf{0.48} &  \textit{0.97} &  \underline{0.66} \\
 HeartSt &    54 &              1.81 &     \textbf{0.56} &  \textit{2.06} &  \underline{1.50} \\
 BreastC &    58 &  \underline{1.27} &     \textbf{0.42} &  \textit{1.71} &              1.39 \\
 HeartHu &    60 &  \underline{0.93} &     \textbf{0.56} &  \textit{1.55} &              1.04 \\
 HeartCl &    62 &              1.69 &     \textbf{0.41} &  \textit{1.96} &  \underline{1.33} \\
 BcWisco &    70 &     \textit{2.02} &     \textbf{0.76} &           1.96 &  \underline{1.14} \\
 Diabete &    77 &  \underline{0.55} &     \textbf{0.23} &  \textit{1.04} &              0.58 \\
 TicTacT &    96 &  \underline{1.60} &     \textbf{0.21} &           2.32 &     \textit{2.47} \\
 Balance &   126 &     \textbf{0.06} &  \underline{0.15} &           0.71 &     \textit{1.18} \\
   Yeast &   149 &              0.54 &     \textbf{0.14} &  \textit{0.78} &  \underline{0.45} \\
 Abalone &   418 &              1.20 &     \textbf{0.78} &  \textit{1.49} &  \underline{0.93} \\
 PageBlo &   548 &     \textit{1.67} &     \textbf{0.38} &           1.17 &  \underline{0.49} \\
 Shuttle &  5800 &     \textit{2.02} &     \textbf{1.05} &           1.84 &  \underline{1.14} \\
   Adult &  6513 &  \underline{1.34} &     \textbf{1.11} &  \textit{1.59} &              1.36 \\
\end{tabular}
\end{center}
\end{table}

### 2017-03-31 intro.tex
proofreading + images etc.

\section{INTRODUCTION}
\label{sec:intro}
Bayesian networks~\cite{Pear88} are popular models for presenting
multivariate statistical dependencies that may have been induced by
underlying causal mechanisms.  Techniques for learning the structure
of Bayesian networks from the observational data has therefore been
used for many tasks such as discovering cell signaling pathways from
the protein activity data~\cite{bn4sigpath02}, revealing the business
process structures~\cite{bn4bpmining} from the transaction logs and
modeling brain region connectivity using fMRI
data~\cite{bn4brainconnect}.

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
identifying the appropriate structure and/or predictive
performance~\cite{cosco.pgm08a}. Traditionally, the most popular
criterion has been the marginal likelihood (usually called BDeu for
reasons explained later), but recent studies~\cite{cosco.uai07,Steck08}
have found this criterion to be very sensitive to hyper parameters
and to yield undesirably complex models for small sample sizes.

The information theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion,  but its exact calculation is likely to be
prohibitively demanding. In 2008, Silander et al. introduced a
hyper-parameter free, NML inspired criterion called a factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without sensitivity problems.  However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, a quotient NML (qNML) that yields simpler models without
sacrificing predictive accuracy. Furthermore, unlike the fNML, the
qNML gives equal scores to the structures that encode the same
independence statements. Like other common model selection criteria,
the qNML is also consistent.

We will next briefly introduce Bayesian networks and then review the
BDeu and fNML criteria and introduce the qNML criterion.  We will also
summarize the results for 20 datasets to back up our claim of qNML
yielding parsimonious models with good predictive capabilities. The
experiments with artificial data demonstrate the capability to give
high scores for the data generating model with relatively small sample
sizes.

>>>>

\section{INTRODUCTION}
\label{sec:intro}
Bayesian networks~\cite{Pear88} are popular models for presenting
multivariate statistical dependencies that may have been induced by
underlying causal mechanisms.  Techniques for learning the structure
of Bayesian networks from observational data have therefore been used
for many tasks such as discovering cell signaling pathways from
protein activity data~\cite{bn4sigpath02}, revealing the business
process structures~\cite{bn4bpmining} from transaction logs and
modeling brain-region connectivity using fMRI
data~\cite{bn4brainconnect}.

Learning the structure of statistical dependencies can be seen as a
model-selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
identifying the appropriate structure and/or predictive
performance~\cite{cosco.pgm08a}. Traditionally, the most popular
criterion has been the marginal likelihood (usually called BDeu for
reasons explained later), but studies~\cite{cosco.uai07,Steck08}
have found this criterion to be very sensitive to hyper parameters
and to yield undesirably complex models for small sample sizes.

The information theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion, but its exact calculation is likely to be
prohibitively expensive. In 2008, Silander et al. introduced a
hyperparameter free, NML inspired criterion called the factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without such sensitivity problems. However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, the \textit{quotient NML} (qNML) that yields simpler models
without sacrificing predictive accuracy. Furthermore, unlike the fNML,
the qNML gives equal scores to structures that encode the same
independence statements. Like other common model selection criteria,
qNML is also consistent.

We next briefly introduce Bayesian networks and review the BDeu and
fNML criteria and then introduce the qNML criterion.  We also
summarize the results for 20 datasets to back up our claim that qNML
yields parsimonious models with good predictive capabilities. The
experiments with artificial data demonstrate the capability to give
high scores for the data generating model with relatively small sample
sizes.
